---
title: "Outbreak Response Scenario Modeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

# Overview

This vignette walks through how to set up, run, and compare outbreak response (OBR) experiments using the `MRTransmissionModel` package. By the end you should be able to:

1. Read and interpret the `setup` object (country demography and vaccination history)
2. Understand every arguments in `EX.Country.part2.OR()`
3. Define a baseline scenario and one or more OBR scenarios
4. Compare scenarios and interpret the results
5. Decide how far into the future to project
6. Know when to use deterministic vs. stochastic mode

---

# 1. Load the package

```{r load}
library(MRTransmissionModel)
source("R/setClasses.R")        # new OBR classes (not yet in installed package)
source("R/run.R")               # new OBR run method
source("R/EX.Country.part2.OR.R")  # new helper function
```

---

# 2. The setup object

`setupCountry.Nov2023()` pulls together all the country-specific inputs you need: demography, vaccination coverage history, and SIA timing. Everything downstream flows from this object.

```{r setup-object}
setup <- setupCountry.Nov2023(country = "Zambia")
names(setup)
```

## 2.1 Demography

### Crude birth rate over time

`cbr.1950.2100` is the crude birth rate (births per 1000 population per year) for every year from 1950 to 2100. The model uses this to update the birth rate at each time step.

```{r cbr-plot}
years.demog <- 1950:2100
plot(years.demog, setup$cbr.1950.2100,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Year", ylab = "Crude birth rate (per 1000)",
     main = "Zambia crude birth rate 1950–2100")
abline(v = c(1980, 2025), lty = 2, col = "gray50")
```

### Total population over time

`pop.total.1950.2100` is the total population for each year. This is used to rescale the simulated population at specified time points so it does not drift from known values.

```{r pop-plot}
plot(years.demog, setup$pop.total.1950.2100 / 1e6,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Year", ylab = "Population (millions)",
     main = "Zambia total population 1950–2100")
abline(v = c(1980, 2025), lty = 2, col = "gray50")
```

## 2.2 Vaccination coverage

Coverage vectors all run from 1980 to 2100. Years beyond the last data year (2022) are filled forward using the mean of 2018, 2019, and 2022 (COVID-robust average).

```{r coverage-plot}
years.vac <- 1980:2100

plot(years.vac, setup$MCV1.coverage.1980to2100,
     type = "l", lwd = 2, col = "steelblue",
     ylim = c(0, 1), xlab = "Year", ylab = "Coverage",
     main = "Zambia vaccination coverage 1980–2100")
lines(years.vac, setup$MCV2.coverage.1980to2100, lwd = 2, col = "darkorange")
lines(years.vac, setup$measlesSIA.coverage.1980to2100, lwd = 2, col = "forestgreen",
      type = "h")  # vertical lines — SIAs are single-year events
legend("topleft",
       legend = c("MCV1 (routine dose 1)", "MCV2 (routine dose 2)", "SIA"),
       col = c("steelblue", "darkorange", "forestgreen"),
       lwd = 2, bty = "n")
```

### SIA age targets

```{r sia-ages}
sia.years <- years.vac[setup$measlesSIA.coverage.1980to2100 > 0]
sia.cov   <- setup$measlesSIA.coverage.1980to2100[setup$measlesSIA.coverage.1980to2100 > 0]
sia.amin  <- setup$age.min.sia.measles[setup$measlesSIA.coverage.1980to2100 > 0]
sia.amax  <- setup$age.max.sia.measles[setup$measlesSIA.coverage.1980to2100 > 0]

data.frame(year     = sia.years,
           coverage = round(sia.cov, 2),
           age.min  = sia.amin,
           age.max  = sia.amax)
```

## 2.3 Vaccine success by age

`pvacsuccess()` converts a vaccine efficacy curve into age-specific probabilities that a vaccine dose results in seroconversion. The Boulianne curve peaks around 12–15 months and falls off at younger ages due to maternal antibody interference.

This matters for OBR because campaigns targeting very young infants will have lower effective coverage than campaigns targeting older children, even if the nominal coverage is the same.

```{r vsucc-plot}
age.classes <- c(1:240, seq(252, 1212, 12))  # months
vsucc.obj   <- pvacsuccess(age.classes, get.boulianne.vsucc())

plot(vsucc.obj@ages, vsucc.obj@prob.vsucc,
     type = "l", lwd = 2, col = "steelblue",
     xlim = c(0, 159),
     xlab = "Age (months)", ylab = "Probability of vaccine success",
     main = "Vaccine success probability by age (Boulianne curve)")
abline(v = c(9, 12, 15, 24), lty = 2, col = "gray60")
text(9, 0.1, "9m\n(MR1 min)", cex = 0.7, adj = 0)
text(15, 0.1, "15m\n(MR2 min)", cex = 0.7, adj = 0)
```

## 2.4 MCV1 age-at-vaccination distribution

`get.MR1cdf.survival()` builds the cumulative distribution of ages at which MCV1 is actually received, accounting for the fact that children do not all receive it on the day they turn 9 months. The distribution is parameterized from country-specific data.

```{r vcdf-plot}
vcdf.MR1 <- get.MR1cdf.survival(uncode = setup$uncode, 1, 24)
vcdf.MR2 <- get.vcdf.normal(25, 36)

par(mfrow = c(1, 2))
plot(vcdf.MR1@ages, vcdf.MR1@cdf,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Age (months)", ylab = "Cumulative fraction vaccinated",
     main = "MCV1 age-at-vaccination CDF")
plot(vcdf.MR2@ages, vcdf.MR2@cdf,
     type = "l", lwd = 2, col = "darkorange",
     xlab = "Age (months)", ylab = "Cumulative fraction vaccinated",
     main = "MCV2 age-at-vaccination CDF")
par(mfrow = c(1, 1))
```

---

# 3. Step 1 — `EX.Country.part1()`: running out the transients

`EX.Country.part1()` spins the model up from a disease-free equilibrium for 20 years so the age structure, maternal immunity, and infection dynamics stabilize before the vaccination scenario begins. The result (`EXt0`) is the shared starting point for all scenarios — run it once and reuse it.

```{r part1, cache=TRUE}
year  <- 1980

EXt0 <- EX.Country.part1(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  age.classes                    = c(1:60, seq(72, 1212, 12)), #c(1:240, seq(252, 1212, 12)),
  maternal.decay.rt              = 0.45,
  exponent                       = 0.97,
  frequency.dep                  = TRUE,
  yr.births.per.1000.acrossyears = rep(setup$cbr.1950.2100[year - 1950 + 1], 20),
  intro.rate                     = 1 / 24 / 320,
  tot.pop                        = NULL,
  targeted.intro                 = FALSE,
  R0                             = 14,
  t.max                          = 20,
  get.births                     = setup$get.births.here,
  seasonal.amp                   = 0.15,
  flat.WAIFW                     = FALSE,
  country.specific.WAIFW         = TRUE,
  vynnycky.waifw                 = FALSE,
  vynnycky.waifw.betas           = c(2, 1),
  asdr.object                    = setup$asdr.object,
  year                           = year,
  use_montagu_demog              = FALSE
)
```

---

# 4. Step 2 — `EX.Country.part2.OR()`: running an Outbreak Response (OBR) scenario

`EX.Country.part2.OR()` takes the equilibrated state from `EXt0` and runs the model forward under a specified vaccination + OBR scenario. We walk through every argument for a single scenario, then build comparisons below.

## 4.0 How the OBR arguments work together

The OBR arguments fall into three groups.

**Timing** — `or.confirmation.delay + or.response.delay = or.total.delay`. At step `t`, the trigger looks backward at cases from steps `[t - or.total.delay - or.trigger.window + 1, t - or.total.delay]`. If the metric reaches `or.n.confirmations.target`, a pending state is set. The campaign fires `or.response.delay` steps later. During the response window, case age distributions accumulate to inform campaign targeting.

| Argument | Default | Description |
|---|---|---|
| `or.confirmation.delay` | required | Steps from end of surveillance window to trigger recognition (confirmation lag) |
| `or.response.delay` | required | Steps from trigger to campaign delivery; case age distribution accumulates during this window |
| `or.trigger.window` | required | Number of steps summed for the trigger metric |

**Trigger** — `or.trigger.mode` controls what is summed over the surveillance window. OBR triggers when `metric >= or.n.confirmations.target`.

| Argument | Default | Description |
|---|---|---|
| `or.trigger.mode` | `"I_scaled"` | `"I_scaled"`: metric = sum(I × reporting rate); `"confirmed"`: metric = sum of estimated confirmed cases from observational model |
| `or.reporting.rate` | `1` | Scales true infections toward observed scale; scalar or per-age vector |
| `or.n.confirmations.target` | `5` | Threshold: OBR triggers when metric ≥ this value; also the per-step testing stop rule in `"confirmed"` mode |
| `or.trigger.age.lower` | `NA` | Lower age bound (months) for surveillance; NA = all ages |
| `or.trigger.age.upper` | `NA` | Upper age bound (months) for surveillance; NA = all ages |
| `or.non.meas.cases.by.age.month` | empty matrix | Age × 12-month matrix of non-measles suspected cases; required for `"confirmed"` mode |
| `or.Se` | `NA` | Diagnostic sensitivity; required for `"confirmed"` mode |
| `or.Sp` | `NA` | Diagnostic specificity; required for `"confirmed"` mode |
| `or.start.timestep` | `1` | Earliest step at which OBR can trigger |
| `or.min.interval` | required | Minimum steps between successive triggers |

**Response campaign** — age targeting is either fixed (`or.vacc.age.upper`) or derived from the cumulative CDF of the case age distribution accumulated during the response window (`or.vacc.agedist.percentile`). Set `or.vacc.coverage = 0` to disable OBR without changing any other argument.

| Argument | Default | Description |
|---|---|---|
| `or.vacc.agedist.percentile` | `NA` | Percentile of the cumulative case age CDF used as upper vaccination bound; used when `or.vacc.age.upper` is NA |
| `or.vacc.age.lower` | `0` | Vaccination floor (months); CDF always counts from age 0 regardless |
| `or.vacc.age.upper` | `NA` | Fixed upper vaccination age (months); if specified, overrides `or.vacc.agedist.percentile` |
| `or.vacc.coverage` | required | Campaign coverage in [0,1]; set to 0 to disable OBR |

**Within-step timing** — at each step: trigger check (looks backward) → campaign delivery if response delay elapsed → state update → case age accumulation for pending response window. In `"confirmed"` mode the per-step confirmed case count is stored after the state update, so future trigger evaluations see the post-campaign case trajectory.

**The observational model runs inline in the loop at every timestep** (when `or.Se`, `or.Sp`, and `or.non.meas.cases.by.age.month` are provided). It serves two roles simultaneously:

1. *Trigger metric*: if `or.trigger.mode == "confirmed"`, the per-step confirmed case count (summed across the trigger age group) is derived from the inline obs model matrices and stored in `confirmed.trigger[]` for use in future trigger evaluations.
2. *Response window accumulation*: age-specific cases are drawn from the inline obs model matrices and accumulated into `pending.case.by.age` — using confirmed cases (TP + FP_test + TP_clinical + FP_clinical) when `or.trigger.mode == "confirmed"`, or suspected cases (I×r + non-measles) when `or.trigger.mode == "I_scaled"`. At campaign delivery, the CDF of this vector determines the upper vaccination age bound.

The six 6-category age × timestep output matrices are stored as slots in the result object and are available immediately after `run()` completes: `result@result@obs.TP`, `result@result@obs.FN_test`, `result@result@obs.FP_test`, `result@result@obs.TN`, `result@result@obs.TP_clinical`, `result@result@obs.FP_clinical`.

## 4.1 Argument walkthrough — one complete scenario

```{r part2-single, cache=TRUE}
t.max <- 45

result.obr <- EX.Country.part2.OR(

  # --- Country and time horizon ---
  uncode                         = setup$uncode,
  year                           = year,
  # Calendar year the simulation starts from (must match part1).

  t.max                          = t.max,
  # Years to simulate forward from year. Here 1980 + 45 = 2025.
  # All coverage vectors must be at least this long.

  EXt0                           = EXt0,
  # The equilibrated starting state from EX.Country.part1().

  # --- Time step and population ---

  generation.time                = 0.5,
  # Length of one time step in months (must match part1 — determines step size throughout).

  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  # Every 10 years, rescale the simulated population to match the WPP total.
  # Prevents the simulated population drifting from reality over long runs.
  # pop.time is in years relative to the start of the simulation.

  is.stochastic                  = FALSE,
  # FALSE = deterministic (fast, reproducible, no extinction).
  # TRUE  = stochastic (run multiple replicates, allows extinction).
  # See Section 6 for discussion.

  rescale.WAIFW                  = FALSE,
  # If TRUE, re-scales the WAIFW matrix to match R0 at the start of part2.
  # Usually FALSE — the scaling was already done in part1.

  # --- Demography ---

  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  # One crude birth rate per year of the simulation (length = t.max + 1).
  # Here we use the actual WPP trajectory — births decline as Zambia develops.

  asdr.object                    = setup$asdr.object,
  # Age-specific death rates. Used to update the survival matrix each year.

  # --- Routine vaccination (MCV1 and MCV2) ---

  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  # One value per year — the proportion of the target age group receiving
  # each routine dose. Drawn from WUENIC estimates.

  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  # Age window (months) for MCV1 delivery. Children aged 9-24 months are eligible.

  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  # Age window for MCV2. 25-36 months.

  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  # CDF of ages at which each dose is actually received, within the age window.
  # MCV1 uses the country-specific survival-based distribution.
  # MCV2 uses a normal distribution centered at 15-21 months within its window.

  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  # Age-specific probability that a dose leads to seroconversion (see plot above).

  # --- Scheduled SIA campaigns ---

  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  # Coverage and age targets for pre-planned SIA campaigns, by year.
  # Zero in years without a campaign.

  sia.timing.in.year             = 3 / 12,
  # Fraction of the year at which the SIA fires (March = 3/12).

  # --- Dose correlation ---

  MR1MR2correlation              = TRUE,
  # TRUE = children who receive MCV1 are more likely to also receive MCV2.
  # Realistic — both doses tend to go to the same children (the "easy to reach").

  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  # Correlation between routine doses and SIA doses.
  # FALSE (= 0) = independent. Positive values mean the same children are
  # reached by both routine and SIA, reducing the net additional coverage.
  # Negative values mean SIAs preferentially reach unvaccinated children.

  # --- Introduction rate ---

  intro.rate                     = 1 / 24 / 320,
  # Rate at which new infections are seeded per age class per time step.
  # Can differ from part1. Lower values reduce background noise in the trigger metric;
  # higher values ensure the model does not go extinct between epidemics.

  # --- OBR trigger definition ---

  or.confirmation.delay          = 2,
  # Time steps from end of surveillance window to case confirmation (lab + reporting).
  # 2 steps = 1 month.

  or.response.delay              = 2,
  # Time steps from confirmation to campaign delivery (mobilization + logistics).
  # 2 steps = 1 month. or.total.delay = or.confirmation.delay + or.response.delay = 4.

  or.trigger.window              = 2,
  # Number of time steps over which cumulative I is summed for the trigger metric.
  # 2 steps = 1 month of summed infectious prevalence.
  # Shorter window = more reactive (responds to recent spikes).
  # Longer window = more smoothed (filters out noise, responds to sustained outbreaks).

  or.trigger.age.lower           = 6,
  or.trigger.age.upper           = 179,
  # Surveillance is restricted to children aged 6 months - 14 years.
  # NA for both = watch all age classes.
  # Focusing surveillance on children reflects where measles is detected first.

  # --- OBR response campaign ---

  or.vacc.age.lower              = 6,
  or.vacc.age.upper              = 59,
  # Campaign targets children aged 6 months - 4 years.
  # OBR campaigns typically target children because they drive transmission.
  # Lower bound of 6 months reflects minimum vaccine-eligible age.

  or.vacc.coverage               = 0.85,
  # Nominal campaign coverage. Applied as independent doses on top of whatever
  # routine/SIA vaccination is already scheduled at the same time step:
  #   effective_prob = p_routine + p_obr - p_routine * p_obr
  # So OBR reaches some already-vaccinated children — net gain is less than 0.85.

  or.min.interval                = 48,
  # Minimum time steps between successive OBR triggers = 2 years.
  # Prevents the model from triggering back-to-back campaigns during a prolonged
  # outbreak that stays above threshold.

  or.start.timestep              = (2020 - year) * 24 + 1
  # OBR cannot trigger before time step (2020-1980)*24+1 = 961.
  # Use this to represent: OBR program not yet established, political will,
  # or a period when you want to measure "what if OBR had been available from 2020."
  # Formula: (target_year - start_year) * steps_per_year + 1
)
```

---

# 5. Building and comparing scenarios

## 5.1 The baseline question

Before adding OBR, you need a baseline to compare against. The right baseline depends on your research question:

| Baseline | When to use |
|----------|-------------|
| No vaccination at all | Measuring total vaccination impact (rarely what you want for OBR) |
| Routine only (MCV1 + MCV2) | Isolating the marginal contribution of SIA + OBR |
| Routine + SIA | Isolating the marginal contribution of OBR alone |
| Routine + SIA + OBR (low coverage) | Comparing OBR coverage levels |

For most OBR questions, **routine + SIA without OBR** is the right baseline. That isolates the effect of the reactive campaign.


## 5.2 Running all four baseline scenarios

```{r sc1, cache=TRUE}
# Scenario 1: No vaccination (OBR disabled via or.vacc.coverage = 0)
sc1 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = rep(0, t.max),
  time.specific.MR2cov           = rep(0, t.max),
  time.specific.SIAcov           = rep(0, t.max),
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = rep(0,  t.max),
  time.specific.max.age.SIA      = rep(0,  t.max),
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.confirmation.delay          = 2,
  or.response.delay              = 2,
  or.trigger.window              = 2,
  or.trigger.age.lower           = 0,
  or.trigger.age.upper           = 168,
  or.vacc.age.lower              = 6,
  or.vacc.age.upper              = 168,
  or.vacc.coverage               = 0,
  or.min.interval                = 48,
  or.start.timestep              = (2020 - year) * 24 + 1
)
```


```{r sc2, cache=TRUE}
# Scenario 2: Routine vaccination only (no SIA, no OBR)
sc2 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = rep(0, t.max),
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = rep(0,  t.max),
  time.specific.max.age.SIA      = rep(0,  t.max),
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.confirmation.delay          = 2,
  or.response.delay              = 2,
  or.trigger.window              = 2,
  or.trigger.age.lower           = 0,
  or.trigger.age.upper           = 168,
  or.vacc.age.lower              = 6,
  or.vacc.age.upper              = 168,
  or.vacc.coverage               = 0,
  or.min.interval                = 48,
  or.start.timestep              = (2020 - year) * 24 + 1
)
```

```{r sc3, cache=TRUE}
# Scenario 3: Routine + SIA (no OBR) — the main baseline for OBR comparison
sc3 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max]*0.3,
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.confirmation.delay          = 2,
  or.response.delay              = 2,
  or.trigger.window              = 2,
  or.trigger.age.lower           = 0,
  or.trigger.age.upper           = 168,
  or.vacc.age.lower              = 6,
  or.vacc.age.upper              = 168,
  or.vacc.coverage               = 0,
  or.min.interval                = 48,
  or.start.timestep              = (2020 - year) * 24 + 1
)
```

## 5.2b Understanding the trigger threshold

The trigger fires when the metric over the surveillance window reaches `or.n.confirmations.target` (default 5). This is the same value that controls when routine testing stops per time step — operationally, the same 5-confirmed-case rule that ends the diagnostic protocol also triggers the OBR campaign.

In `"I_scaled"` mode, `or.reporting.rate` must be calibrated so that scaled infections are on the same scale as confirmed cases (i.e., reaching 5 scaled infections corresponds to reaching 5 confirmed cases in the field). In `"confirmed"` mode, the metric is directly the estimated total confirmed case count from the observational model, so no additional calibration of `or.n.confirmations.target` is needed.

A practical approach: run a scenario (no OBR) and plot the trigger metric over time to verify that `or.n.confirmations.target = 5` fires at plausible outbreak moments.

```{r threshold-exploration}
rc     <- sc3@result@.Data
i.inds <- sc3@result@i.inds
age.cl <- sc3@result@age.class
n      <- ncol(rc)

# Restrict to 0-14 years (0-168 months) — the trigger age group
trigger.age.pos <- which(age.cl > 0 & age.cl <= 168)
trigger.i.inds  <- i.inds[trigger.age.pos]

# Sum I across the trigger age group at each time step
I.timestep <- colSums(rc[trigger.i.inds, ])
t.vec      <- year + (seq_len(n) - 1) * 0.5 / 12

# Rolling sum over w time steps (NA for the first w-1 steps)
rolling.sum <- function(x, w) {
  sapply(seq_along(x), function(t) if (t < w) NA else sum(x[(t - w + 1):t]))
}

plot(t.vec, I.timestep,
     type = "l", col = "gray70", lwd = 1,
     xlab = "Year", ylab = "Sum of I (ages 0-14)",
     main = "Incidence at different smoothing windows")
lines(t.vec, rolling.sum(I.timestep, 2), col = "steelblue",   lwd = 1.5)
lines(t.vec, rolling.sum(I.timestep, 4), col = "darkorange",  lwd = 1.5)
lines(t.vec, rolling.sum(I.timestep, 8), col = "forestgreen", lwd = 1.5)
legend("topright",
       legend = c("1 step (2 wks)", "2 steps (1 mo)", "4 steps (2 mo)", "8 steps (4 mo)"),
       col    = c("gray70", "steelblue", "darkorange", "forestgreen"),
       lwd    = c(1, 1.5, 1.5, 1.5), bty = "n")
```

### Key learning point: model I vs. reported cases

The y-axis here is the **sum of I from the model** — meaning true infectious individuals. In `"I_scaled"` mode, the trigger metric is `sum(I × reporting_rate)` — scaled infections. The reporting rate `r` (set via `or.reporting.rate`) converts true infections to roughly the scale of reported/confirmed cases, so that reaching `or.n.confirmations.target = 5` corresponds to 5 confirmed cases in the field.

In practice, only a fraction φ ("phi") of true infections ever become a reported case. A child has to:

1. develop symptoms severe enough to seek care
2. be recognized as a suspected measles case by a clinician
3. have a specimen collected and tested (if confirmation is required)

Estimates of φ for measles in high-burden settings often range from 1% to 10%, meaning 90–99% of infections are never counted. The `or.reporting.rate` parameter captures this fraction — calibrating it so that `sum(I × r) >= 5` fires when the surveillance system would genuinely observe 5 confirmed cases.

In `"confirmed"` mode, the observational model handles this scaling automatically via Se, Sp, and the non-measles background, so `or.reporting.rate` need only capture the pre-testing reporting fraction (care-seeking × clinical recognition).

Back to the other scenarios.

```{r sc4, cache=TRUE}
# Scenario 4: Routine + SIA + OBR
sc4 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max]*0.3,
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.trigger.mode                = "I_scaled",
  or.reporting.rate              = 0.01,
  or.n.confirmations.target      = 5,
  or.confirmation.delay          = 2,
  or.response.delay              = 8,
  or.trigger.window              = 2,
  or.trigger.age.lower           = 0,
  or.trigger.age.upper           = 168,
  or.vacc.age.lower              = 6,
  or.vacc.age.upper              = 168,
  or.vacc.agedist.percentile     = NA_real_,
  or.vacc.coverage               = 0.8,
  or.min.interval                = 48,
  or.start.timestep              = (2005 - year) * 24 + 1
)
```

## 5.3 Comparing results

```{r compare-plot}
get.annual.I <- function(result, year.start, t.max) {
  rc     <- result@result@.Data
  i.inds <- result@result@i.inds
  n.steps.per.year <- 24
  sapply(1:t.max, function(yr) {
    cols <- ((yr - 1) * n.steps.per.year + 1):(yr * n.steps.per.year)
    sum(rc[i.inds, cols])
  })
}

years.sim <- year + 1:t.max - 0.5

I.sc1 <- get.annual.I(sc1, year, t.max)
I.sc2 <- get.annual.I(sc2, year, t.max)
I.sc3 <- get.annual.I(sc3, year, t.max)
I.sc4 <- get.annual.I(sc4, year, t.max)

plot(years.sim, I.sc1, type = "l", lwd = 2, col = "gray50",
     xlab = "Year", ylab = "Annual sum of I",
     main = "Measles burden by scenario",
     ylim = range(c(I.sc1, I.sc2, I.sc3, I.sc4)))
lines(years.sim, I.sc2, lwd = 2, col = "steelblue")
lines(years.sim, I.sc3, lwd = 2, col = "darkorange")
lines(years.sim, I.sc4, lwd = 2, col = "forestgreen")

# Mark OBR campaign years
obr.steps <- which(sc4@result@or.times == 1)
obr.yrs   <- year + (obr.steps - 1) * 0.5 / 12
abline(v = obr.yrs, lty = 3, col = "forestgreen", lwd = 0.8)

legend("topright",
       legend = c("No vaccination", "Routine only", "Routine + SIA",
                  "Routine + SIA + OBR", "OBR campaign"),
       col    = c("gray50", "steelblue", "darkorange", "forestgreen", "forestgreen"),
       lty    = c(1, 1, 1, 1, 3), lwd = 2, bty = "n")
```

### Cumulative burden table

```{r burden-table}
data.frame(
  Scenario         = c("No vaccination", "Routine only",
                       "Routine + SIA", "Routine + SIA + OBR"),
  Cumulative_I     = round(c(sum(I.sc1), sum(I.sc2), sum(I.sc3), sum(I.sc4))),
  Pct_reduction_vs_sc3 = round(100 * (1 - c(sum(I.sc1), sum(I.sc2),
                                              sum(I.sc3), sum(I.sc4)) / sum(I.sc3)), 1)
)
```

### When did OBR campaigns fire?

```{r obr-times}
obr.steps <- which(sc4@result@or.times == 1)
obr.years <- year + (obr.steps - 1) * 0.5 / 12
cat("OBR campaigns fired at calendar years:\n")
print(round(obr.years, 2))
cat("\nTotal OBR campaigns:", length(obr.years), "\n")
```

---

# 6. How far into the future to model

The right time horizon depends on your question:

| Question | Suggested horizon |
|----------|-------------------|
| Evaluate OBR impact on a known recent outbreak | 5–10 years around the outbreak |
| Compare OBR scenarios over a policy-relevant period | 10–20 years (e.g., 2025–2045) |
| Evaluate long-run trajectory under different strategies | 30–50 years |

**Practical considerations:**

- **Burn-in**: The first ~5 years of a forward projection are heavily influenced by the starting state. If you start in 1980, the 1980–1985 period reflects transient dynamics, not steady state under the vaccination program.
- **Coverage extrapolation**: Coverage beyond 2022 is assumed constant (held at the 2018–2022 average). Long projections become increasingly speculative.
- **OBR availability**: Use `or.start.timestep` to prevent OBR from firing in years when the program did not exist.
- **Population rescaling**: For runs beyond ~20 years, population rescaling (`pop.rescale`) is important to prevent drift.
- MORE.... what do you think they are?

A sensible approach for OBR analysis: start in 1980 (to properly burn in the vaccination program history), but focus comparisons on 2020 onward (when OBR is activated via `or.start.timestep`).

---

# 7. Deterministic vs. stochastic

## 7.1 What changes with `is.stochastic = TRUE`

In deterministic mode (`is.stochastic = FALSE`), transitions between compartments are exact expected values. In stochastic mode, each transition is drawn from a multinomial distribution — individuals are randomly assigned to outcomes.

**Consequences:**

| Property | Deterministic | Stochastic |
|----------|--------------|------------|
| Speed | Fast | Slower (~10× per replicate) |
| Reproducibility | Always identical | Varies by run |
| Extinction | Never — I can be fractional | Possible — I hits 0 |
| Epidemic timing | Fixed | Varies across replicates |
| Appropriate for | Scenario comparison, threshold calibration | Uncertainty quantification, extinction probability |

## 7.2 When to use each

Use **deterministic** when:
- Exploring the parameter space (threshold values, coverage levels)
- Confirming OBR campaigns fire at the right times
- Generating a single expected trajectory

Use **stochastic** when:
- You need uncertainty intervals around projected burden
- You want the probability that OBR prevents an outbreak entirely (via extinction)
- Your population is small (stochastic effects matter more at low N)

## 7.3 Running stochastic replicates

```{r stochastic-demo, eval=FALSE}
n.reps <- 50
I.stoch <- matrix(NA, nrow = n.reps, ncol = t.max)

for (i in 1:n.reps) {
  res <- EX.Country.part2.OR(
    # ... same arguments as sc4 above ...
    is.stochastic = TRUE
  )
  I.stoch[i, ] <- get.annual.I(res, year, t.max)
}

# Median and 95% interval
I.med   <- apply(I.stoch, 2, median)
I.lo    <- apply(I.stoch, 2, quantile, 0.025)
I.hi    <- apply(I.stoch, 2, quantile, 0.975)

plot(years.sim, I.med, type = "l", lwd = 2, col = "forestgreen",
     xlab = "Year", ylab = "Annual sum of I",
     main = "Stochastic uncertainty — Routine + SIA + OBR",
     ylim = range(c(I.lo, I.hi)))
polygon(c(years.sim, rev(years.sim)), c(I.hi, rev(I.lo)),
        col = adjustcolor("forestgreen", 0.2), border = NA)
lines(years.sim, I.sc4, lwd = 2, lty = 2, col = "black")  # deterministic
legend("topright",
       legend = c("Stochastic median", "95% interval", "Deterministic"),
       col = c("forestgreen", adjustcolor("forestgreen", 0.2), "black"),
       lty = c(1, NA, 2), lwd = c(2, NA, 2),
       fill = c(NA, adjustcolor("forestgreen", 0.2), NA),
       border = c(NA, "forestgreen", NA), bty = "n")
```

---

# 8. Try for yourself.

1. **Threshold sensitivity**: How does changing `or.n.confirmations.target` from 5 to 10 change the number of OBR campaigns that fire and the total burden reduction?

2. **Window length**: A longer `or.trigger.window` (e.g., 4 steps = 2 months, or 24 steps = 1 year) makes the trigger smoother — it responds to sustained outbreaks rather than single-step spikes. Does this help or hurt? Under what outbreak dynamics would the 1-month window over-trigger?

3. **Delay**: If detection and response lag increases from 2 months (`or.total.delay = 4`) to 6 months (`or.total.delay = 12`), how much impact does OBR lose?

4. **Age targeting**: Should the OBR campaign target the same age group as the surveillance trigger? What are the tradeoffs of a narrow vs. wide age target?

5. **Start year**: What happens if you set `or.start.timestep = 1` (OBR available from 1980)? Is this realistic? Does it change the long-run burden substantially?



Amy List:
- update setup to include wpp2024
- update setup to include 2024 SIA
- update setup to include Abraham's MCV1 and MCV2 age-specific estimates
- update boulliane to move over one age bin
- if should rescale pop
