---
title: "Outbreak Response Scenario Modeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "..")
```

# Overview

This vignette walks through how to set up, run, and compare outbreak response (OBR) experiments using the `MRTransmissionModel` package. By the end you should be able to:

1. Read and interpret the `setup` object (country demography and vaccination history)
2. Understand every arguments in `EX.Country.part2.OR()`
3. Define a baseline scenario and one or more OBR scenarios
4. Compare scenarios and interpret the results
5. Decide how far into the future to project
6. Know when to use deterministic vs. stochastic mode

---

# 1. Load the package

```{r load}
library(MRTransmissionModel)
source("R/setClasses.R")        # new OBR classes (not yet in installed package)
source("R/run.R")               # new OBR run method
source("R/EX.Country.part2.OR.R")  # new helper function
```

---

# 2. The setup object

`setupCountry.Nov2023()` pulls together all the country-specific inputs you need: demography, vaccination coverage history, and SIA timing. Everything downstream flows from this object.

```{r setup-object}
setup <- setupCountry.Nov2023(country = "Zambia")
names(setup)
```

## 2.1 Demography

### Crude birth rate over time

`cbr.1950.2100` is the crude birth rate (births per 1000 population per year) for every year from 1950 to 2100. The model uses this to update the birth rate at each time step.

```{r cbr-plot}
years.demog <- 1950:2100
plot(years.demog, setup$cbr.1950.2100,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Year", ylab = "Crude birth rate (per 1000)",
     main = "Zambia crude birth rate 1950–2100")
abline(v = c(1980, 2025), lty = 2, col = "gray50")
```

### Total population over time

`pop.total.1950.2100` is the total population for each year. This is used to rescale the simulated population at specified time points so it does not drift from known values.

```{r pop-plot}
plot(years.demog, setup$pop.total.1950.2100 / 1e6,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Year", ylab = "Population (millions)",
     main = "Zambia total population 1950–2100")
abline(v = c(1980, 2025), lty = 2, col = "gray50")
```

## 2.2 Vaccination coverage

Coverage vectors all run from 1980 to 2100. Years beyond the last data year (2022) are filled forward using the mean of 2018, 2019, and 2022 (COVID-robust average).

```{r coverage-plot}
years.vac <- 1980:2100

plot(years.vac, setup$MCV1.coverage.1980to2100,
     type = "l", lwd = 2, col = "steelblue",
     ylim = c(0, 1), xlab = "Year", ylab = "Coverage",
     main = "Zambia vaccination coverage 1980–2100")
lines(years.vac, setup$MCV2.coverage.1980to2100, lwd = 2, col = "darkorange")
lines(years.vac, setup$measlesSIA.coverage.1980to2100, lwd = 2, col = "forestgreen",
      type = "h")  # vertical lines — SIAs are single-year events
legend("topleft",
       legend = c("MCV1 (routine dose 1)", "MCV2 (routine dose 2)", "SIA"),
       col = c("steelblue", "darkorange", "forestgreen"),
       lwd = 2, bty = "n")
```

### SIA age targets

```{r sia-ages}
sia.years <- years.vac[setup$measlesSIA.coverage.1980to2100 > 0]
sia.cov   <- setup$measlesSIA.coverage.1980to2100[setup$measlesSIA.coverage.1980to2100 > 0]
sia.amin  <- setup$age.min.sia.measles[setup$measlesSIA.coverage.1980to2100 > 0]
sia.amax  <- setup$age.max.sia.measles[setup$measlesSIA.coverage.1980to2100 > 0]

data.frame(year     = sia.years,
           coverage = round(sia.cov, 2),
           age.min  = sia.amin,
           age.max  = sia.amax)
```

## 2.3 Vaccine success by age

`pvacsuccess()` converts a vaccine efficacy curve into age-specific probabilities that a vaccine dose results in seroconversion. The Boulianne curve peaks around 12–15 months and falls off at younger ages due to maternal antibody interference.

This matters for OBR because campaigns targeting very young infants will have lower effective coverage than campaigns targeting older children, even if the nominal coverage is the same.

```{r vsucc-plot}
age.classes <- c(1:240, seq(252, 1212, 12))  # months
vsucc.obj   <- pvacsuccess(age.classes, get.boulianne.vsucc())

plot(vsucc.obj@ages, vsucc.obj@prob.vsucc,
     type = "l", lwd = 2, col = "steelblue",
     xlim = c(0, 159),
     xlab = "Age (months)", ylab = "Probability of vaccine success",
     main = "Vaccine success probability by age (Boulianne curve)")
abline(v = c(9, 12, 15, 24), lty = 2, col = "gray60")
text(9, 0.1, "9m\n(MR1 min)", cex = 0.7, adj = 0)
text(15, 0.1, "15m\n(MR2 min)", cex = 0.7, adj = 0)
```

## 2.4 MCV1 age-at-vaccination distribution

`get.MR1cdf.survival()` builds the cumulative distribution of ages at which MCV1 is actually received, accounting for the fact that children do not all receive it on the day they turn 9 months. The distribution is parameterized from country-specific data.

```{r vcdf-plot}
vcdf.MR1 <- get.MR1cdf.survival(uncode = setup$uncode, 1, 24)
vcdf.MR2 <- get.vcdf.normal(25, 36)

par(mfrow = c(1, 2))
plot(vcdf.MR1@ages, vcdf.MR1@cdf,
     type = "l", lwd = 2, col = "steelblue",
     xlab = "Age (months)", ylab = "Cumulative fraction vaccinated",
     main = "MCV1 age-at-vaccination CDF")
plot(vcdf.MR2@ages, vcdf.MR2@cdf,
     type = "l", lwd = 2, col = "darkorange",
     xlab = "Age (months)", ylab = "Cumulative fraction vaccinated",
     main = "MCV2 age-at-vaccination CDF")
par(mfrow = c(1, 1))
```

---

# 3. Step 1 — `EX.Country.part1()`: running out the transients

`EX.Country.part1()` spins the model up from a disease-free equilibrium for 20 years so the age structure, maternal immunity, and infection dynamics stabilize before the vaccination scenario begins. The result (`EXt0`) is the shared starting point for all scenarios — run it once and reuse it.

```{r part1, cache=TRUE}
year  <- 1980

EXt0 <- EX.Country.part1(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  age.classes                    = c(1:60, seq(72, 1212, 12)), #c(1:240, seq(252, 1212, 12)),
  maternal.decay.rt              = 0.45,
  exponent                       = 0.97,
  frequency.dep                  = TRUE,
  yr.births.per.1000.acrossyears = rep(setup$cbr.1950.2100[year - 1950 + 1], 20),
  intro.rate                     = 1 / 24 / 320,
  tot.pop                        = NULL,
  targeted.intro                 = FALSE,
  R0                             = 14,
  t.max                          = 20,
  get.births                     = setup$get.births.here,
  seasonal.amp                   = 0.15,
  flat.WAIFW                     = FALSE,
  country.specific.WAIFW         = TRUE,
  vynnycky.waifw                 = FALSE,
  vynnycky.waifw.betas           = c(2, 1),
  asdr.object                    = setup$asdr.object,
  year                           = year,
  use_montagu_demog              = FALSE
)
```

---

# 4. Step 2 — `EX.Country.part2.OR()`: running an Outbreak Response (OBR) scenario

`EX.Country.part2.OR()` takes the equilibrated state from `EXt0` and runs the model forward under a specified vaccination + OBR scenario. We walk through every argument for a single scenario, then build comparisons below.

## 4.0 How the OBR arguments work together

The OBR arguments fall into three groups.

**Timing** — `or.confirmation.delay + or.response.delay = or.total.delay`. At step `t`, the trigger looks backward at IgM-confirmed cases (obs.TP) from steps `[t - or.total.delay - or.trigger.window + 1, t - or.total.delay]`. If the sum reaches `or.n.confirmations.target`, a pending state is set. The campaign fires `or.response.delay` steps later. During the response window, case age distributions accumulate to inform campaign targeting.

| Argument | Default | Description |
|---|---|---|
| `or.confirmation.delay` | required | Steps from end of surveillance window to trigger recognition. **RDT** (point-of-care rapid test): ≈1 step (∼2 weeks — results available near-immediately). **EIA/ELISA** (lab-based): ≈2 steps (∼1 month for specimen transport, lab processing, and reporting) |
| `or.response.delay` | required | Steps from trigger to campaign delivery; case age distribution accumulates during this window |
| `or.trigger.window` | required | Number of steps over which obs.TP is summed for the trigger metric |

**Trigger** — the trigger always fires on IgM-confirmed cases (obs.TP) summed over the surveillance window and age group. The observational model (Se, Sp, non-measles background) runs at every time step.

| Argument | Default | Description |
|---|---|---|
| `or.reporting.rate` | `1` | Pre-testing fraction: care-seeking × clinical recognition; converts true I to suspected cases entering surveillance |
| `or.non.meas.cases.by.age.month` | required | Age × 12-month matrix of non-measles rash-febrile cases; rows = model age classes, columns = calendar months |
| `or.Se` | required | Diagnostic test sensitivity ∈ [0,1] |
| `or.Sp` | required | Diagnostic test specificity ∈ [0,1] |
| `or.n.confirmations.target` | `5` | Trigger threshold (sum of obs.TP over window must reach this) and per-step testing stop rule |
| `or.trigger.window.age.lower` | `NA` | Lower age bound (months) for surveillance; NA = all ages |
| `or.trigger.window.age.upper` | `NA` | Upper age bound (months) for surveillance; NA = all ages |
| `or.start.timestep` | `1` | Earliest step at which OBR can trigger |
| `or.min.interval` | required | Minimum steps between successive triggers |

**Response campaign** — the upper vaccination age bound is always derived from the CDF of case ages accumulated during the response window. `stop_testing_at_trigger` controls which cases go into that CDF. Set `or.vacc.coverage = 0` to disable OBR.

| Argument | Default | Description |
|---|---|---|
| `or.vacc.agedist.percentile` | required | Percentile (0–1) of the cumulative case age CDF used as the upper vaccination age bound |
| `stop_testing_at_trigger` | `FALSE` | Controls which cases build the campaign age-targeting CDF. **`FALSE` (RDT)**: CDF built from obs.TP (IgM-confirmed cases); testing continues through the response window — feasible when RDT results are available at point-of-care. **`TRUE` (EIA/ELISA)**: CDF built from all suspected cases (true.reported + non-measles background; reporting process only) — appropriate when lab turnaround is long enough that field teams rely on clinical epi rather than confirmed results for age targeting |
| `or.vacc.coverage` | required | Campaign coverage in [0,1]; set to 0 to disable OBR |

**Within-step timing** — at each step: trigger check (looks backward at obs.TP) → campaign delivery if response delay elapsed → state update → observational model fills obs.TP[,t] → accumulate case ages for pending response window.

**The observational model runs at every time step** and produces six age × timestep matrices stored on the result object: `result@result@obs.TP`, `result@result@obs.FN_test`, `result@result@obs.FP_test`, `result@result@obs.TN`, `result@result@obs.TP_clinical`, `result@result@obs.FP_clinical`.

## 4.1 Argument walkthrough — one complete scenario

```{r part2-single, cache=TRUE}
t.max <- 45

result.obr <- EX.Country.part2.OR(

  # --- Country and time horizon ---
  uncode                         = setup$uncode,
  year                           = year,
  # Calendar year the simulation starts from (must match part1).

  t.max                          = t.max,
  # Years to simulate forward from year. Here 1980 + 45 = 2025.
  # All coverage vectors must be at least this long.

  EXt0                           = EXt0,
  # The equilibrated starting state from EX.Country.part1().

  # --- Time step and population ---

  generation.time                = 0.5,
  # Length of one time step in months (must match part1 — determines step size throughout).

  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  # Every 10 years, rescale the simulated population to match the WPP total.
  # Prevents the simulated population drifting from reality over long runs.
  # pop.time is in years relative to the start of the simulation.

  is.stochastic                  = FALSE,
  # FALSE = deterministic (fast, reproducible, no extinction).
  # TRUE  = stochastic (run multiple replicates, allows extinction).
  # See Section 6 for discussion.

  rescale.WAIFW                  = FALSE,
  # If TRUE, re-scales the WAIFW matrix to match R0 at the start of part2.
  # Usually FALSE — the scaling was already done in part1.

  # --- Demography ---

  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  # One crude birth rate per year of the simulation (length = t.max + 1).
  # Here we use the actual WPP trajectory — births decline as Zambia develops.

  asdr.object                    = setup$asdr.object,
  # Age-specific death rates. Used to update the survival matrix each year.

  # --- Routine vaccination (MCV1 and MCV2) ---

  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  # One value per year — the proportion of the target age group receiving
  # each routine dose. Drawn from WUENIC estimates.

  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  # Age window (months) for MCV1 delivery. Children aged 9-24 months are eligible.

  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  # Age window for MCV2. 25-36 months.

  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  # CDF of ages at which each dose is actually received, within the age window.
  # MCV1 uses the country-specific survival-based distribution.
  # MCV2 uses a normal distribution centered at 15-21 months within its window.

  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  # Age-specific probability that a dose leads to seroconversion (see plot above).

  # --- Scheduled SIA campaigns ---

  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  # Coverage and age targets for pre-planned SIA campaigns, by year.
  # Zero in years without a campaign.

  sia.timing.in.year             = 3 / 12,
  # Fraction of the year at which the SIA fires (March = 3/12).

  # --- Dose correlation ---

  MR1MR2correlation              = TRUE,
  # TRUE = children who receive MCV1 are more likely to also receive MCV2.
  # Realistic — both doses tend to go to the same children (the "easy to reach").

  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  # Correlation between routine doses and SIA doses.
  # FALSE (= 0) = independent. Positive values mean the same children are
  # reached by both routine and SIA, reducing the net additional coverage.
  # Negative values mean SIAs preferentially reach unvaccinated children.

  # --- Introduction rate ---

  intro.rate                     = 1 / 24 / 320,
  # Rate at which new infections are seeded per age class per time step.
  # Can differ from part1. Lower values reduce background noise in the trigger metric;
  # higher values ensure the model does not go extinct between epidemics.

  # --- Observational model ---
  # Required by all OBR scenarios. Converts true infections into the surveillance
  # case counts that drive the trigger and campaign age targeting.

  or.reporting.rate              = 0.10,
  # Pre-testing fraction: care-seeking × clinical recognition.
  # ~10% of true infections present to a health facility as a suspected case.
  # This is before any diagnostic test is done.

  or.non.meas.cases.by.age.month = outer(
    0.4 * exp(-EXt0@trans@age.class / 48),          # age-declining background rate
    c(1.15, 1.15, 1.10, 1.05, 1.00, 0.90,           # Jan-Jun seasonal multipliers
      0.85, 0.85, 0.90, 0.95, 1.05, 1.15)),         # Jul-Dec seasonal multipliers
  # Non-measles suspected cases per time step by age class (rows) and calendar month (columns).
  # The entries are counts of non-measles rash-febrile illness reaching the surveillance system.
  # Here: exponentially declining with age (peak in infants), 20% higher in Zambia rainy season.

  or.Se                          = 0.95,
  # IgM test sensitivity. Standard measles value from the literature.

  or.Sp                          = 0.97,
  # IgM test specificity.

  # --- OBR trigger ---

  or.n.confirmations.target      = 5,
  # Trigger threshold: OBR fires when the sum of obs.TP (IgM-confirmed measles)
  # over the trigger age group and window reaches this value.
  # Also controls the per-step testing budget: test enough suspected cases to expect
  # this many positives, then stop.

  or.confirmation.delay          = 1,
  # Steps from end of surveillance window to trigger recognition.
  # RDT (point-of-care): 1 step ≈ 2 weeks — results available near-immediately.
  # EIA/ELISA alternative: 2 steps ≈ 1 month for specimen transport + lab processing.

  or.response.delay              = 6,
  # Steps from trigger recognition to campaign delivery (mobilization + logistics).
  # 6 steps = 3 months. or.total.delay = 1 + 6 = 7.

  or.trigger.window              = 4,
  # Number of steps over which obs.TP is summed for the trigger metric.
  # 4 steps = 2 months. Shorter = more reactive; longer = smoother signal.

  or.trigger.window.age.lower    = 6,
  or.trigger.window.age.upper    = 179,
  # Surveillance restricted to children 6 months – 14 years.
  # NA for both = watch all age classes.

  # --- OBR response campaign ---

  or.vacc.coverage               = 0.85,
  # Nominal campaign coverage. Applied on top of any routine/SIA dose at the same step:
  #   effective_prob = p_routine + p_obr - p_routine * p_obr
  # So OBR reaches some already-vaccinated children — net gain is less than 0.85.

  or.vacc.agedist.percentile     = 0.95,
  # The upper vaccination age bound is the 95th percentile of the cumulative case age CDF
  # built from cases accumulated during the response window.
  # A value of 0.95 means: vaccinate all ages from the youngest age class up to the age
  # below which 95% of outbreak cases occurred.

  stop_testing_at_trigger        = FALSE,
  # FALSE (RDT scenario): case age CDF is built from obs.TP — IgM-confirmed cases only.
  #   RDT results are available at point-of-care, so testing continues through the response
  #   window and the CDF reflects confirmed measles.
  # TRUE (EIA/ELISA scenario): CDF is built from all suspected cases (true.reported +
  #   non-measles background). Use when lab turnaround means field teams rely on clinical
  #   epi for age targeting rather than waiting for confirmed results.

  or.min.interval                = 48,
  # Minimum steps between successive OBR triggers = 2 years.
  # Prevents back-to-back campaigns during a prolonged outbreak above threshold.

  or.start.timestep              = (2020 - year) * 24 + 1
  # OBR cannot trigger before this step — here 2020.
  # Formula: (target_year - start_year) * steps_per_year + 1
)
```

---

# 5. Building and comparing scenarios

## 5.1 The baseline question

Before adding OBR, you need a baseline to compare against. The right baseline depends on your research question:

| Baseline | When to use |
|----------|-------------|
| No vaccination at all | Measuring total vaccination impact (rarely what you want for OBR) |
| Routine only (MCV1 + MCV2) | Isolating the marginal contribution of SIA + OBR |
| Routine + SIA | Isolating the marginal contribution of OBR alone |
| Routine + SIA + OBR (low coverage) | Comparing OBR coverage levels |

For most OBR questions, **routine + SIA without OBR** is the right baseline. That isolates the effect of the reactive campaign.


## 5.2 Running all four baseline scenarios

```{r shared-obs-setup}
# Shared observational model inputs used by all four scenarios.
# Build once; pass to each EX.Country.part2.OR() call.
age.bg.rate <- 0.4 * exp(-EXt0@trans@age.class / 48)
month.mult  <- c(1.15, 1.15, 1.10, 1.05, 1.00, 0.90,
                 0.85, 0.85, 0.90, 0.95, 1.05, 1.15)
non.meas.bg <- outer(age.bg.rate, month.mult)
```

```{r sc1, cache=TRUE}
# Scenario 1: No vaccination (OBR disabled via or.vacc.coverage = 0)
sc1 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = rep(0, t.max),
  time.specific.MR2cov           = rep(0, t.max),
  time.specific.SIAcov           = rep(0, t.max),
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = rep(0,  t.max),
  time.specific.max.age.SIA      = rep(0,  t.max),
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.reporting.rate              = 0.10,
  or.non.meas.cases.by.age.month = non.meas.bg,
  or.Se                          = 0.95,
  or.Sp                          = 0.97,
  or.n.confirmations.target      = 5,
  or.confirmation.delay          = 2,
  or.response.delay              = 2,
  or.trigger.window              = 2,
  or.trigger.window.age.lower    = 0,
  or.trigger.window.age.upper    = 168,
  or.vacc.coverage               = 0,
  or.vacc.agedist.percentile     = 0.95,
  stop_testing_at_trigger        = FALSE,
  or.min.interval                = 48,
  or.start.timestep              = (2020 - year) * 24 + 1
)
```


```{r sc2, cache=TRUE}
# Scenario 2: Routine vaccination only (no SIA, no OBR)
sc2 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = rep(0, t.max),
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = rep(0,  t.max),
  time.specific.max.age.SIA      = rep(0,  t.max),
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.reporting.rate              = 0.10,
  or.non.meas.cases.by.age.month = non.meas.bg,
  or.Se                          = 0.95,
  or.Sp                          = 0.97,
  or.n.confirmations.target      = 5,
  or.confirmation.delay          = 2,
  or.response.delay              = 2,
  or.trigger.window              = 2,
  or.trigger.window.age.lower    = 0,
  or.trigger.window.age.upper    = 168,
  or.vacc.coverage               = 0,
  or.vacc.agedist.percentile     = 0.95,
  stop_testing_at_trigger        = FALSE,
  or.min.interval                = 48,
  or.start.timestep              = (2020 - year) * 24 + 1
)
```

```{r sc3, cache=TRUE}
# Scenario 3: Routine + SIA (no OBR) — the main baseline for OBR comparison
sc3 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max]*0.3,
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.reporting.rate              = 0.10,
  or.non.meas.cases.by.age.month = non.meas.bg,
  or.Se                          = 0.95,
  or.Sp                          = 0.97,
  or.n.confirmations.target      = 5,
  or.confirmation.delay          = 2,
  or.response.delay              = 2,
  or.trigger.window              = 2,
  or.trigger.window.age.lower    = 0,
  or.trigger.window.age.upper    = 168,
  or.vacc.coverage               = 0,
  or.vacc.agedist.percentile     = 0.95,
  stop_testing_at_trigger        = FALSE,
  or.min.interval                = 48,
  or.start.timestep              = (2020 - year) * 24 + 1
)
```

## 5.2b Understanding the trigger threshold

The trigger fires when the sum of `obs.TP` (IgM-confirmed measles cases) over the surveillance window and age group reaches `or.n.confirmations.target` (default 5). This is the same value that controls the per-step testing budget: the model tests just enough suspected cases to expect 5 IgM positives, then stops.

`or.n.confirmations.target = 5` operationally represents the outbreak detection threshold used in the field — a cluster of 5 confirmed cases is the signal that prompts an OBR response.

A practical approach: run sc3 (no OBR) and plot `obs.TP` over time to verify the threshold fires at plausible outbreak moments.

```{r threshold-exploration}
rc.tp  <- sc3@result@obs.TP
age.cl <- sc3@result@age.class
n      <- ncol(rc.tp)

# Restrict to 0-14 years — the trigger age group
trigger.age.pos <- which(age.cl > 0 & age.cl <= 168)

# Sum obs.TP across the trigger age group at each time step
TP.timestep <- colSums(rc.tp[trigger.age.pos, ])
t.vec       <- year + (seq_len(n) - 1) * 0.5 / 12

# Rolling sum over w time steps
rolling.sum <- function(x, w) {
  sapply(seq_along(x), function(t) if (t < w) NA else sum(x[(t - w + 1):t]))
}

plot(t.vec, TP.timestep,
     type = "l", col = "gray70", lwd = 1,
     xlab = "Year", ylab = "obs.TP (IgM-confirmed, ages 0-14)",
     main = "Confirmed measles cases at different smoothing windows")
lines(t.vec, rolling.sum(TP.timestep, 2), col = "steelblue",   lwd = 1.5)
lines(t.vec, rolling.sum(TP.timestep, 4), col = "darkorange",  lwd = 1.5)
lines(t.vec, rolling.sum(TP.timestep, 8), col = "forestgreen", lwd = 1.5)
abline(h = 5, lty = 2, col = "red", lwd = 1)
legend("topright",
       legend = c("1 step (2 wks)", "2 steps (1 mo)", "4 steps (2 mo)", "8 steps (4 mo)",
                  "threshold = 5"),
       col    = c("gray70", "steelblue", "darkorange", "forestgreen", "red"),
       lty    = c(1, 1, 1, 1, 2),
       lwd    = c(1, 1.5, 1.5, 1.5, 1), bty = "n")
```

### Key learning point: the two-stage observation process

The trigger operates on `obs.TP`, not on raw model infections. Two parameters govern how true infections become `obs.TP`:

- **`or.reporting.rate`** (pre-testing): what fraction of true infections reach a health facility as a suspected case (care-seeking × clinical recognition). Typically 1–10% in high-burden settings.
- **`or.Se`** (diagnostic): given a true measles suspected case that gets tested, what is the probability the IgM test comes back positive.

Together they determine `obs.TP ≈ I × or.reporting.rate × or.Se`. Calibrate `or.reporting.rate` so that `obs.TP` over the trigger window reaches 5 at plausible outbreak moments — check against the plot above.

Back to the other scenarios.

```{r sc4, cache=TRUE}
# Scenario 4: Routine + SIA + RDT OBR (age targets from IgM-confirmed cases)
# RDT (rapid diagnostic test): near-zero lab delay — results available at point-of-care.
# Testing continues through the response window, so the campaign age CDF is built from
# confirmed measles cases (obs.TP). or.confirmation.delay = 1 (~2-week turnaround).
sc4 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max]*0.3,
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.reporting.rate              = 0.10,
  or.non.meas.cases.by.age.month = non.meas.bg,
  or.Se                          = 0.95,
  or.Sp                          = 0.97,
  or.n.confirmations.target      = 5,
  or.confirmation.delay          = 1,           # RDT: ~2-week point-of-care turnaround
  or.response.delay              = 6,           # 3-month mobilization lag
  or.trigger.window              = 4,           # 2-month lookback window
  or.trigger.window.age.lower    = 0,
  or.trigger.window.age.upper    = 168,
  or.vacc.coverage               = 0.85,
  or.vacc.agedist.percentile     = 0.95,
  stop_testing_at_trigger        = FALSE,       # RDT: age CDF from confirmed cases (obs.TP)
  or.min.interval                = 48,
  or.start.timestep              = (2005 - year) * 24 + 1
)
```

```{r sc5, cache=TRUE}
# Scenario 5: Routine + SIA + EIA OBR (age targets from suspected cases)
# EIA/ELISA (enzyme immunoassay): ~1-month lab turnaround means confirmed results are
# not available before the campaign launches. stop_testing_at_trigger = TRUE uses
# clinical suspected cases for age targeting rather than waiting for lab confirmation.
# or.confirmation.delay = 2 (~1 month for specimen transport + lab processing + reporting).
sc5 <- EX.Country.part2.OR(
  uncode                         = setup$uncode,
  generation.time                = 0.5,
  pop.rescale                    = setup$pop.total.1950.2100[
                                     seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                       = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                  = FALSE,
  t.max                          = t.max,
  rescale.WAIFW                  = FALSE,
  yr.births.per.1000.acrossyears = setup$cbr.1950.2100[
                                     (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                    = setup$asdr.object,
  year                           = year,
  EXt0                           = EXt0,
  time.specific.MR1cov           = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov           = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov           = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max]*0.3,
  time.specific.min.age.MR1      = rep(9,  t.max),
  time.specific.max.age.MR1      = rep(24, t.max),
  time.specific.min.age.MR2      = rep(25, t.max),
  time.specific.max.age.MR2      = rep(36, t.max),
  time.specific.min.age.SIA      = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA      = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  obj.vcdf.MR1                   = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                   = get.vcdf.normal(25, 36),
  obj.prob.vsucc                 = pvacsuccess(EXt0@trans@age.class,
                                               get.boulianne.vsucc()),
  sia.timing.in.year             = 3 / 12,
  MR1MR2correlation              = TRUE,
  MR1SIAcorrelation              = FALSE,
  MR2SIAcorrelation              = FALSE,
  intro.rate                     = 1 / 24 / 320,
  or.reporting.rate              = 0.10,
  or.non.meas.cases.by.age.month = non.meas.bg,
  or.Se                          = 0.95,
  or.Sp                          = 0.97,
  or.n.confirmations.target      = 5,
  or.confirmation.delay          = 2,           # EIA: ~1-month lab turnaround
  or.response.delay              = 6,           # 3-month mobilization lag
  or.trigger.window              = 4,           # 2-month lookback window
  or.trigger.window.age.lower    = 0,
  or.trigger.window.age.upper    = 168,
  or.vacc.coverage               = 0.85,
  or.vacc.agedist.percentile     = 0.95,
  stop_testing_at_trigger        = TRUE,        # EIA: age CDF from suspected cases
  or.min.interval                = 48,
  or.start.timestep              = (2005 - year) * 24 + 1
)
```

## 5.3 Comparing results

```{r compare-plot}
get.annual.I <- function(result, year.start, t.max) {
  rc     <- result@result@.Data
  i.inds <- result@result@i.inds
  n.steps.per.year <- 24
  sapply(1:t.max, function(yr) {
    cols <- ((yr - 1) * n.steps.per.year + 1):(yr * n.steps.per.year)
    sum(rc[i.inds, cols])
  })
}

years.sim <- year + 1:t.max - 0.5

I.sc1 <- get.annual.I(sc1, year, t.max)
I.sc2 <- get.annual.I(sc2, year, t.max)
I.sc3 <- get.annual.I(sc3, year, t.max)
I.sc4 <- get.annual.I(sc4, year, t.max)
I.sc5 <- get.annual.I(sc5, year, t.max)

plot(years.sim, I.sc1, type = "l", lwd = 2, col = "gray50",
     xlab = "Year", ylab = "Annual sum of I",
     main = "Measles burden by scenario",
     ylim = range(c(I.sc1, I.sc2, I.sc3, I.sc4, I.sc5)))
lines(years.sim, I.sc2, lwd = 2, col = "steelblue")
lines(years.sim, I.sc3, lwd = 2, col = "darkorange")
lines(years.sim, I.sc4, lwd = 2, col = "forestgreen")
lines(years.sim, I.sc5, lwd = 2, col = "purple")

# Mark OBR campaign years (sc4 = RDT, sc5 = EIA)
obr.steps.sc4 <- which(sc4@result@or.times == 1)
obr.steps.sc5 <- which(sc5@result@or.times == 1)
abline(v = year + (obr.steps.sc4 - 1) * 0.5 / 12, lty = 3, col = "forestgreen", lwd = 0.8)
abline(v = year + (obr.steps.sc5 - 1) * 0.5 / 12, lty = 3, col = "purple",      lwd = 0.8)

legend("topright",
       legend = c("No vaccination", "Routine only", "Routine + SIA",
                  "Routine + SIA + RDT OBR", "Routine + SIA + EIA OBR"),
       col    = c("gray50", "steelblue", "darkorange", "forestgreen", "purple"),
       lty    = c(1, 1, 1, 1, 1), lwd = 2, bty = "n")
```

### Cumulative burden table

```{r burden-table}
data.frame(
  Scenario         = c("No vaccination", "Routine only", "Routine + SIA",
                       "Routine + SIA + RDT OBR", "Routine + SIA + EIA OBR"),
  Cumulative_I     = round(c(sum(I.sc1), sum(I.sc2), sum(I.sc3),
                               sum(I.sc4), sum(I.sc5))),
  Pct_reduction_vs_sc3 = round(100 * (1 - c(sum(I.sc1), sum(I.sc2), sum(I.sc3),
                                              sum(I.sc4), sum(I.sc5)) / sum(I.sc3)), 1)
)
```

### When did OBR campaigns fire?

```{r obr-times}
obr.steps.sc4 <- which(sc4@result@or.times == 1)
obr.years.sc4 <- year + (obr.steps.sc4 - 1) * 0.5 / 12
cat("RDT OBR campaigns (sc4) fired at calendar years:\n")
print(round(obr.years.sc4, 2))
cat("Total:", length(obr.years.sc4), "\n\n")

obr.steps.sc5 <- which(sc5@result@or.times == 1)
obr.years.sc5 <- year + (obr.steps.sc5 - 1) * 0.5 / 12
cat("EIA OBR campaigns (sc5) fired at calendar years:\n")
print(round(obr.years.sc5, 2))
cat("Total:", length(obr.years.sc5), "\n")
```

---

# 6. How far into the future to model

The right time horizon depends on your question:

| Question | Suggested horizon |
|----------|-------------------|
| Evaluate OBR impact on a known recent outbreak | 5–10 years around the outbreak |
| Compare OBR scenarios over a policy-relevant period | 10–20 years (e.g., 2025–2045) |
| Evaluate long-run trajectory under different strategies | 30–50 years |

**Practical considerations:**

- **Burn-in**: The first ~5 years of a forward projection are heavily influenced by the starting state. If you start in 1980, the 1980–1985 period reflects transient dynamics, not steady state under the vaccination program.
- **Coverage extrapolation**: Coverage beyond 2022 is assumed constant (held at the 2018–2022 average). Long projections become increasingly speculative.
- **OBR availability**: Use `or.start.timestep` to prevent OBR from firing in years when the program did not exist.
- **Population rescaling**: For runs beyond ~20 years, population rescaling (`pop.rescale`) is important to prevent drift.
- MORE.... what do you think they are?

A sensible approach for OBR analysis: start in 1980 (to properly burn in the vaccination program history), but focus comparisons on 2020 onward (when OBR is activated via `or.start.timestep`).

---

# 7. Deterministic vs. stochastic

## 7.1 What changes with `is.stochastic = TRUE`

In deterministic mode (`is.stochastic = FALSE`), transitions between compartments are exact expected values. In stochastic mode, each transition is drawn from a multinomial distribution — individuals are randomly assigned to outcomes.

**Consequences:**

| Property | Deterministic | Stochastic |
|----------|--------------|------------|
| Speed | Fast | Slower (~10× per replicate) |
| Reproducibility | Always identical | Varies by run |
| Extinction | Never — I can be fractional | Possible — I hits 0 |
| Epidemic timing | Fixed | Varies across replicates |
| Appropriate for | Scenario comparison, threshold calibration | Uncertainty quantification, extinction probability |

## 7.2 When to use each

Use **deterministic** when:
- Exploring the parameter space (threshold values, coverage levels)
- Confirming OBR campaigns fire at the right times
- Generating a single expected trajectory

Use **stochastic** when:
- You need uncertainty intervals around projected burden
- You want the probability that OBR prevents an outbreak entirely (via extinction)
- Your population is small (stochastic effects matter more at low N)

## 7.3 Running stochastic replicates

```{r stochastic-demo, eval=FALSE}
n.reps <- 50
I.stoch <- matrix(NA, nrow = n.reps, ncol = t.max)

for (i in 1:n.reps) {
  res <- EX.Country.part2.OR(
    # ... same arguments as sc4 above ...
    is.stochastic = TRUE
  )
  I.stoch[i, ] <- get.annual.I(res, year, t.max)
}

# Median and 95% interval
I.med   <- apply(I.stoch, 2, median)
I.lo    <- apply(I.stoch, 2, quantile, 0.025)
I.hi    <- apply(I.stoch, 2, quantile, 0.975)

plot(years.sim, I.med, type = "l", lwd = 2, col = "forestgreen",
     xlab = "Year", ylab = "Annual sum of I",
     main = "Stochastic uncertainty — Routine + SIA + OBR",
     ylim = range(c(I.lo, I.hi)))
polygon(c(years.sim, rev(years.sim)), c(I.hi, rev(I.lo)),
        col = adjustcolor("forestgreen", 0.2), border = NA)
lines(years.sim, I.sc4, lwd = 2, lty = 2, col = "black")  # deterministic
legend("topright",
       legend = c("Stochastic median", "95% interval", "Deterministic"),
       col = c("forestgreen", adjustcolor("forestgreen", 0.2), "black"),
       lty = c(1, NA, 2), lwd = c(2, NA, 2),
       fill = c(NA, adjustcolor("forestgreen", 0.2), NA),
       border = c(NA, "forestgreen", NA), bty = "n")
```

---

# 8. Try for yourself.

1. **Threshold sensitivity**: How does changing `or.n.confirmations.target` from 5 to 10 change the number of OBR campaigns that fire and the total burden reduction?

2. **Window length**: A longer `or.trigger.window` (e.g., 4 steps = 2 months, or 24 steps = 1 year) makes the trigger smoother — it responds to sustained outbreaks rather than single-step spikes. Does this help or hurt? Under what outbreak dynamics would the 1-month window over-trigger?

3. **Delay**: If detection and response lag increases from 2 months (`or.total.delay = 4`) to 6 months (`or.total.delay = 12`), how much impact does OBR lose?

4. **Age targeting**: Should the OBR campaign target the same age group as the surveillance trigger? What are the tradeoffs of a narrow vs. wide age target?

5. **Start year**: What happens if you set `or.start.timestep = 1` (OBR available from 1980)? Is this realistic? Does it change the long-run burden substantially?



Amy List:
- update setup to include wpp2024
- update setup to include 2024 SIA
- update setup to include Abraham's MCV1 and MCV2 age-specific estimates
- update boulliane to move over one age bin
- if should rescale pop
