---
title: "Getting Started with MRTransmissionModel"
author: "MRTransmissionModel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MRTransmissionModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`MRTransmissionModel` is an age-structured, discrete-time stochastic
transmission model for measles and rubella. The model tracks five
epidemiological stages — Maternally immune (M), Susceptible (S), Infected
(I), Recovered (R), and effectively Vaccinated (V) — across many age classes and 
uses a transition matrix at every time step to move people between those epidemiological classes.

Most users will only ever interact with three wrapper functions, which
should be called **in this order**:

1. `setupCountry.Nov2023()` — pulls demography (UNWPP 2022) and vaccination coverage (WHO/WUENIC, SIA) data for a single country.
2. `EX.Country.part1()` — builds the transmission experiment object and runs out the transients to reach a stable starting state.
3. `EX.Country.part2()` — runs the actual simulation under whatever vaccination scenario you specify.

This vignette walks through each step using **Zambia** as the example
country. The same workflow applies to any country supported by the
underlying WHO and UNPD data files.

```{r}
library(MRTransmissionModel)
```

---

## Step 1: `setupCountry.Nov2023()`

This function gathers everything you need about a country in one call:
demographic projections (1950–2100), routine MCV1/MCV2 and RCV1 coverage
(1980–2100), measles and rubella SIA campaigns, age-specific birth and
death rates, and the inaccessible-to-vaccine proportion.

```{r}
setup <- setupCountry.Nov2023(country = "Zambia")
```

### Argument

- `country` — character string. The country name as recognized by the
  `countrycode` package (e.g. `"Zambia"`, `"Kenya"`, `"Pakistan"`).
  Internally this is converted to both UN code and ISO3 code.

### What you get back

`setup` is a named list. The pieces you'll usually pass on to the next two
functions include:

- `uncode`, `iso3code` — country identifiers.
- `cbr.1950.2100` — crude birth rate per 1000 per year.
- `pop.total.1950.2100`, `pop.age.byageclasses.1950.2100` — total and
  age-structured population.
- `asdr.object` — age-specific death rate object used to build survival.
- `get.births.here` — pre-built births function for the country.
- `MCV1.coverage.1980to2100`, `MCV2.coverage.1980to2100`,
  `RCV1.coverage.1980to2100` — annual routine coverage (proportions).
- `measlesSIA.coverage.1980to2100`, `age.min.sia.measles`,
  `age.max.sia.measles` — measles SIA coverage and age targeting.
- `rubellaSIA.coverage.1980to2100`, `age.min.sia.rubella`,
  `age.max.sia.rubella` — same for rubella SIAs.
- `inaccessible.prop.1980to2100` — proportion of the population assumed
  unreachable by vaccination (derived from DTP1).

---

## Step 2: `EX.Country.part1()`

This sets up the transmission experiment and **runs out the transients**
so that the population is in a realistic equilibrium before the actual
simulation begins. The output (`EXt0`) is the starting point for Step 3.

```{r}
year  <- 1980

EXt0 <- EX.Country.part1(
  uncode                       = setup$uncode,
  generation.time              = 0.5,
  age.classes                  = c(1:240, seq(252, 1212, 12)),
  maternal.decay.rt            = 0.45,
  exponent                     = 0.97,
  frequency.dep                = TRUE,
  yr.births.per.1000.acrossyears = rep(setup$cbr.1950.2100[year - 1950 + 1], 20),
  intro.rate                   = 1 / 24 / 320,
  tot.pop                      = NULL,
  targeted.intro               = FALSE,
  R0                           = 20,
  t.max                        = 20,
  get.births                   = setup$get.births.here,
  seasonal.amp                 = 0.15,
  flat.WAIFW                   = FALSE,
  country.specific.WAIFW       = TRUE,
  vynnycky.waifw               = FALSE,
  vynnycky.waifw.betas         = c(2, 1),
  asdr.object                  = setup$asdr.object,
  year                         = year,
  use_montagu_demog            = FALSE
)

```

### Key arguments

- `uncode` — UN country code (from `setup$uncode`).
- `generation.time` — generation time in **months** (0.5 ≈ 2 weeks, the
  standard for measles/rubella).
- `age.classes` — vector of upper age class boundaries in months. The
  default `c(1:240, seq(252, 1212, 12))` gives monthly resolution from 0–20
  years and yearly resolution from 20–101 years.
- `maternal.decay.rt` — rate of maternal antibody decay.
- `exponent`, `frequency.dep` — mixing assumptions for the force of
  infection.
- `yr.births.per.1000.acrossyears` — vector of crude birth rates (one per
  year of `t.max`). For the transient run it's typical to repeat the start
  year's CBR.
- `intro.rate` — rate at which new infections are introduced (seeds the
  model so it doesn't go extinct).
- `targeted.intro` — `FALSE` spreads introductions across age classes;
  `TRUE` concentrates them.
- `R0` — basic reproduction number used to rescale the WAIFW (contact)
  matrix. Measles is typically 12–18; rubella is typically 5–7.
- `t.max` — number of **years** to run the transients (20 is usually
  enough).
- `get.births` — births function (use `setup$get.births.here`).
- `seasonal.amp` — amplitude of the seasonal cosine in transmission
  (0.15–0.35 is typical).
- `flat.WAIFW`, `country.specific.WAIFW`, `vynnycky.waifw` — choose the
  mixing matrix. `country.specific.WAIFW = TRUE` uses the Prem et al.
  country-specific contact matrix.
- `vynnycky.waifw.betas` — only used if `vynnycky.waifw = TRUE`; two
  betas (young, old).
- `asdr.object` — age-specific death rates (use `setup$asdr.object`).
- `year` — the calendar year whose demography is used as the starting
  point (DFE — disease-free equilibrium).
- `use_montagu_demog` — `TRUE` to use Montagu (VIMC) demography instead of
  UNPD; usually `FALSE`.
- `routine.vac`, `routine.vac.age.index` — optional routine vaccination
  during the transient phase (usually 0).

The function returns an `experiment.updatedemog` object (`EXt0`) that
holds the equilibrated state, transition matrix, contact matrix, and
seasonal/maternal pieces — everything Step 3 needs.

---

## Step 3: `EX.Country.part2()`

This runs the **actual simulation** forward in time under a specified
vaccination scenario. You can call it multiple times with the same `EXt0`
to compare scenarios (e.g., no-vaccine vs. routine + SIA).

### A) No-vaccination scenario

```{r}
t.max <- 45
no_vac <- EX.Country.part2(
  uncode                          = setup$uncode,
  generation.time                 = 0.5,
  pop.rescale                     = setup$pop.total.1950.2100[
                                      seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                        = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                   = FALSE,
  get.births                      = setup$get.births.here,
  t.max                           = t.max,
  rescale.WAIFW                   = FALSE,
  yr.births.per.1000.acrossyears  = setup$cbr.1950.2100[
                                      (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                     = setup$asdr.object,
  year                            = year,
  EXt0                            = EXt0,
  time.specific.MR1cov            = rep(0, t.max),
  time.specific.MR2cov            = rep(0, t.max),
  time.specific.SIAcov            = rep(0, t.max),
  time.specific.min.age.MR1       = rep(0, t.max),
  time.specific.max.age.MR1       = rep(0, t.max),
  time.specific.min.age.MR2       = rep(0, t.max),
  time.specific.max.age.MR2       = rep(0, t.max),
  time.specific.min.age.SIA       = rep(0, t.max),
  time.specific.max.age.SIA       = rep(0, t.max),
  obj.vcdf.MR1                    = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                    = get.vcdf.normal(25, 36),
  obj.prob.vsucc                  = pvacsuccess(c(1:240, seq(252, 1212, 12)),
                                                get.boulianne.vsucc()),
  sia.timing.in.year              = 3 / 12,
  MR1MR2correlation               = TRUE,
  MR1SIAcorrelation               = FALSE,
  MR2SIAcorrelation               = FALSE,
  SIAinacc                        = FALSE,
  prop.inacc                      = rep(0, t.max),
  SIAinefficient                  = FALSE,
  intro.rate                      = 1 / 24 / 320
)

plot(no_vac@result)
```

### B) With-vaccination scenario

Swap the zero coverage vectors for the country's actual coverage and SIA
age targeting from `setup`:

```{r}
with_vac <- EX.Country.part2(
  uncode                          = setup$uncode,
  generation.time                 = 0.5,
  pop.rescale                     = setup$pop.total.1950.2100[
                                      seq(1990, (year + t.max - 10), 10) - 1950 + 1],
  pop.time                        = seq(1990, (year + t.max - 10), 10) - year,
  is.stochastic                   = TRUE,
  get.births                      = setup$get.births.here,
  t.max                           = t.max,
  rescale.WAIFW                   = FALSE,
  yr.births.per.1000.acrossyears  = setup$cbr.1950.2100[
                                      (year - 1950 + 1):((year - 1950 + 1) + t.max)],
  asdr.object                     = setup$asdr.object,
  year                            = year,
  EXt0                            = EXt0,
  time.specific.MR1cov            = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov            = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov            = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.min.age.MR1       = rep(9,  t.max),
  time.specific.max.age.MR1       = rep(24, t.max),
  time.specific.min.age.MR2       = rep(25, t.max),
  time.specific.max.age.MR2       = rep(36, t.max),
  time.specific.min.age.SIA       = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA       = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  obj.vcdf.MR1                    = get.MR1cdf.survival(uncode = setup$uncode, 1, 24),
  obj.vcdf.MR2                    = get.vcdf.normal(25, 36),
  obj.prob.vsucc                  = pvacsuccess(c(1:240, seq(252, 1212, 12)),
                                                get.boulianne.vsucc()),
  sia.timing.in.year              = 3 / 12,
  MR1MR2correlation               = FALSE,
  MR1SIAcorrelation               = FALSE,
  MR2SIAcorrelation               = FALSE,
  SIAinacc                        = FALSE,
  prop.inacc                      = rep(0, t.max),
  SIAinefficient                  = FALSE,
  intro.rate                      = 1 / 24 / 320
)

plot(with_vac@result)
```

### Key arguments

- `uncode`, `generation.time`, `year`, `asdr.object`, `get.births`,
  `yr.births.per.1000.acrossyears` — the same as in Step 2; usually pulled
  straight from `setup`.
- `EXt0` — the experiment object returned by `EX.Country.part1()`.
- `t.max` — number of **years** to simulate forward (e.g. `2022:2100` →
  `t.max = 79`).
- `pop.rescale`, `pop.time` — optional vectors to rescale the simulated
  population to known totals at specific years (keeps long runs from
  drifting).
- `is.stochastic` — `TRUE` for stochastic dynamics (recommended for actual
  simulations); `FALSE` for deterministic.
- `rescale.WAIFW` — usually `FALSE` after Step 2 has already rescaled.
- `time.specific.MR1cov`, `time.specific.MR2cov`, `time.specific.SIAcov` —
  vectors of length `t.max` giving the proportion vaccinated each year by
  routine MR1, routine MR2, and SIAs respectively. Use 0s for a
  no-vaccine counterfactual.
- `time.specific.min.age.MR1` / `max.age.MR1` (and MR2, SIA) — vectors of
  length `t.max` giving the min/max **age in months** targeted each year.
- `obj.vcdf.MR1`, `obj.vcdf.MR2` — cumulative vaccination distributions
  (when in age people actually get the dose). Defaults are reasonable;
  `get.MR1cdf.survival()`, `get.vcdf.normal()`, and `get.vcdf.uniform()`
  are alternatives.
- `obj.prob.vsucc` — probability that a dose successfully immunizes by
  age. `pvacsuccess(... , get.boulianne.vsucc())` is the standard.
- `sia.timing.in.year` — fraction of the year when SIAs occur (e.g.
  `3/12` ≈ April).
- `MR1MR2correlation`, `MR1SIAcorrelation`, `MR2SIAcorrelation` — assume
  the same children get repeatedly missed/reached across doses.
- `SIAinacc`, `prop.inacc` — model a fraction of the population as
  inaccessible to SIAs (use `setup$inaccessible.prop.1980to2100` if you
  want this, subset to your year range).
- `SIAinefficient` — flag SIAs as less than fully efficient.
- `intro.rate` — scalar or vector controlling new infection introductions
  over time.

The return value is a `sim.results.MSIRV` object whose `@result` slot
holds the M/S/I/R/V counts by age and time step. Use `plot()` for a
quick look.

### Digging into `with_vac`

`with_vac` is an S4 `experiment.result` object with two top-level slots:

- `with_vac@experiment.def` — the experiment definition that was run
  (transition matrix, contact matrix, seasonal/maternal pieces, etc.).
  Handy if you want to recover the exact inputs used.
- `with_vac@result` — a `sim.results.MSIRV.update.demog.vaccine.change`
  object (a subclass of `sim.results.MSIRV`, which itself extends a plain
  `matrix`). This is where the simulated state lives.

The `@result` object has several slots worth knowing about:

- `@.Data` — the underlying numeric **matrix** of counts. Rows are
  compartment-by-age-class combinations (all M age classes stacked, then
  all S, then I, then R, then V). Columns are time steps. So
  `dim(with_vac@result@.Data)` is roughly
  `(5 * length(age.classes)) x (t.max / generation.time)`.
- `@m.inds`, `@s.inds`, `@i.inds`, `@r.inds`, `@v.inds` — integer vectors
  giving the **row indices** in `@.Data` that belong to each compartment
  (Maternally immune / Susceptible / Infected / Recovered / Vaccinated).
- `@age.class` — upper age-class boundaries in months (same vector you
  passed to `EX.Country.part1()`).
- `@t` — the vector of simulation time steps (in years, relative to
  `year`).
- `@births.each.timestep`, `@growth.rate.each.timestep`,
  `@MR1.fail.each.timestep`, `@MR2.fail.each.timestep`,
  `@SIA.fail.each.timestep` — bookkeeping vectors recorded during the
  run, useful for tallying births, missed doses, and vaccine failures.

### Using `@.Data` to pull out pieces

Because `@.Data` is just a matrix, the `@*.inds` slots are all you need
to slice out whatever compartment, age range, or time window you want:

```{r}
res <- with_vac@result

# Full compartment time series (sum across all age classes)
S_t <- colSums(res@.Data[res@s.inds, ])   # susceptibles over time
I_t <- colSums(res@.Data[res@i.inds, ])   # infecteds over time
V_t <- colSums(res@.Data[res@v.inds, ])   # effectively vaccinated

# Age-structured snapshot at a single time step (e.g. the last one)
last <- ncol(res@.Data)
I_by_age_last <- res@.Data[res@i.inds, last]
names(I_by_age_last) <- res@age.class     # labelled by upper age in months

# Infecteds in children under 5 years (i.e. age class <= 60 months)
under5 <- which(res@age.class <= 60)
I_u5_t <- colSums(res@.Data[res@i.inds[under5], ])

# Annual incidence: sum infecteds within each calendar year
year_idx  <- floor(res@t)                 # integer year offset
I_by_year <- tapply(colSums(res@.Data[res@i.inds, ]), year_idx, sum)
```

If you'd rather not index by hand, `getCompartmentTotals(with_vac@result)`
collapses `@.Data` to a 5-row matrix (M/S/I/R/V) by time, and
`plot(with_vac@result)` gives the default visual summary.

---

## Putting it all together

A complete minimal run looks like:

```{r}
library(MRTransmissionModel)

setup <- setupCountry.Nov2023(country = "Zambia")
year  <- 2022
t.max <- length(year:2100)

EXt0 <- EX.Country.part1(
  uncode = setup$uncode,
  yr.births.per.1000.acrossyears = rep(setup$cbr.1950.2100[year - 1950 + 1], 20),
  asdr.object = setup$asdr.object,
  get.births  = setup$get.births.here,
  year        = year,
  R0          = 5,
  t.max       = 20,
  country.specific.WAIFW = TRUE
)

sim <- EX.Country.part2(
  uncode = setup$uncode,
  EXt0   = EXt0,
  year   = year,
  t.max  = t.max,
  asdr.object = setup$asdr.object,
  get.births  = setup$get.births.here,
  yr.births.per.1000.acrossyears =
    setup$cbr.1950.2100[(year - 1950 + 1):((year - 1950 + 1) + t.max)],
  time.specific.MR1cov  = setup$MCV1.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.MR2cov  = setup$MCV2.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.SIAcov  = setup$measlesSIA.coverage.1980to2100[(year - 1980 + 1):t.max],
  time.specific.min.age.MR1 = rep(9,  t.max),
  time.specific.max.age.MR1 = rep(24, t.max),
  time.specific.min.age.MR2 = rep(25, t.max),
  time.specific.max.age.MR2 = rep(36, t.max),
  time.specific.min.age.SIA = setup$age.min.sia.measles[(year - 1980 + 1):t.max],
  time.specific.max.age.SIA = setup$age.max.sia.measles[(year - 1980 + 1):t.max],
  is.stochastic = TRUE,
  intro.rate    = 1 / 24 / 320
)

plot(sim@result)
```

That's the full workflow: **set up → run transients → simulate**. Swap
the country, change the coverage vectors, or vary `R0` to explore other
scenarios.

## Where to learn more

- See `?setupCountry.Nov2023`, `?EX.Country.part1`, and
  `?EX.Country.part2` for the full argument lists.
- `auxiliary/runScript_who_un_inputs.R` in the package source has a
  fuller worked example with multiple scenarios.
- The model framework is described in Metcalf et al. (2012), *Theor Popul
  Biol* and *Epidemiol Infect*; see the package README for the full
  citations.
